Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent coevolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable (*CITE*). Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).

>>>>

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent coevolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable [@Hansen2014]. Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).
